# Data Wrangling with dplyr

For this lecture, we will use the built-in `mtcars` dataset and save it as a CSV to demonstrate the import process:

```{r}
# First, let's save mtcars as a CSV file for demonstration
write.csv(mtcars, "mtcars_example.csv", row.names = TRUE)

# Now import it back
car_data = read.csv("mtcars_example.csv")
```

## What is dplyr?

**dplyr** is a powerful R package that provides a consistent set of functions (called "verbs") for manipulating data frames. It is part of the **tidyverse**, a collection of R packages designed for data science.

The key advantages of dplyr include:

-   **Readable syntax**: Functions are named as verbs that describe what they do
-   **Consistent interface**: All functions work the same way
-   **Speed**: Optimized for performance on large datasets
-   **Pipe operator**: Allows chaining multiple operations together

We will also introduce **tidyr**, another tidyverse package that provides functions for reshaping data between wide and long formats.

## Installing and Loading dplyr and tidyr

```{r, eval=FALSE}
# Install packages (only need to do this once)
install.packages("dplyr")
install.packages("tidyr")

# Or install the entire tidyverse (includes both)
install.packages("tidyverse")
```

```{r}
# Load tidyverse (includes dplyr and tidyr)
library(tidyverse)
```

## Core dplyr Verbs
### `rename()`: Rename Columns
The `rename()` function changes the names of columns in a data frame. This is useful when column names are unclear, too long, or inconsistent with your naming conventions.

**Syntax:**

```{r, eval=FALSE}
rename(data, new_name = old_name, ...)
```

**Key arguments:**

-   `new_name = old_name`: The new column name on the left, the existing column name on the right
-   Can rename multiple columns in one call
-   Column names with spaces or special characters need backticks: `` `old name` ``

```{r}
# Rename a single column
head(rename(car_data, miles_per_gallon = mpg), 3)
```

```{r}
# Rename multiple columns
head(rename(car_data, 
            miles_per_gallon = mpg, 
            horsepower = hp,
            weight = wt), 3)
```

```{r}
# Save renamed data to a new variable
car_renamed = rename(car_data, cylinders = cyl, gear_count = gear)
head(car_renamed, 3)
```


### `select()`: Choose Columns

The `select()` function is used to choose which columns (variables) to keep in your data frame. This is useful when you have a dataset with many variables but only need a few for your analysis.

**Syntax:**

```{r, eval=FALSE}
select(data, column1, column2, ...)
```

**Key arguments:**

-   Column names (unquoted)
-   Column indices (numbers)
-   Helper functions: `starts_with()`, `ends_with()`, `contains()`, `matches()`
-   Use `-` to exclude columns

```{r}
# Select specific columns by name
select(car_data, mpg, hp, wt)
```

```{r}
# Select columns 2 through 5
select(car_data, 2:5)
```

```{r}
# Remove columns using minus sign
select(car_data, -X, -vs, -am)
```

```{r}
# Select columns that start with a pattern
select(car_data, starts_with("c"))
```

### `filter()`: Choose Rows

The `filter()` function is used to subset rows based on logical conditions. Only rows where the condition evaluates to `TRUE` are kept. This is equivalent to the `WHERE` clause in SQL.

**Syntax:**

```{r, eval=FALSE}
filter(data, condition1, condition2, ...)
```

**Key arguments:**

-   Logical conditions using comparison operators: `==`, `!=`, `<`, `>`, `<=`, `>=`
-   Combine conditions with `&` (AND) or `,` (also AND), `|` (OR)
-   Use `%in%` to match multiple values
-   Use `!` to negate conditions

```{r}
# Filter rows where mpg > 25
filter(car_data, mpg > 25)
```

```{r}
# Multiple conditions (AND)
filter(car_data, mpg > 20, cyl == 4)
```

```{r}
# Multiple conditions (OR)
filter(car_data, cyl == 4 | cyl == 6)
```

```{r}
# Using %in% for multiple values
filter(car_data, cyl %in% c(4, 6))
```

### `arrange()`: Sort Rows

The `arrange()` function reorders rows based on the values of one or more columns. By default, it sorts in ascending order. Use `desc()` to sort in descending order. This is equivalent to `ORDER BY` in SQL.

**Syntax:**

```{r, eval=FALSE}
arrange(data, column1, column2, ...)
```

**Key arguments:**

-   Column names to sort by (unquoted)
-   Wrap column name in `desc()` for descending order
-   Multiple columns: sorts by first column, then by second within ties, etc.

```{r}
# Sort by mpg (ascending)
head(arrange(car_data, mpg), 5)
```

```{r}
# Sort by mpg (descending)
head(arrange(car_data, desc(mpg)), 5)
```

```{r}
# Sort by multiple columns
head(arrange(car_data, cyl, desc(mpg)), 10)
```

### `mutate()`: Create New Variables

The `mutate()` function creates new columns (variables) or modifies existing ones. The new variables are added to the end of the data frame. You can reference other columns in the same `mutate()` call, including ones you just created.

**Syntax:**

```{r, eval=FALSE}
mutate(data, new_column = expression, ...)
```

**Key arguments:**

-   `new_column = expression`: Name of new column and how to calculate it
-   Can use arithmetic operators: `+`, `-`, `*`, `/`, `^`
-   Can use functions: `log()`, `sqrt()`, `abs()`, `round()`, etc.
-   Can reference existing columns and newly created columns

```{r}
# Create a new variable (kilometers per liter)
car_data_new = mutate(car_data, kpl = mpg * 0.425)
head(select(car_data_new, X, mpg, kpl), 5)
```

```{r}
# Create multiple new variables
car_data_new = mutate(car_data, 
                       hp_per_cyl = hp / cyl,
                       weight_kg = wt * 453.592)
head(select(car_data_new, X, hp, cyl, hp_per_cyl, wt, weight_kg), 5)
```

### `summarize()`: Calculate Summary Statistics

The `summarize()` function (also spelled `summarise()`) collapses a data frame into a single row of summary statistics. It is most powerful when combined with `group_by()` to calculate summaries for each group.

**Syntax:**

```{r, eval=FALSE}
summarize(data, summary_name = summary_function(column), ...)
```

**Key arguments:**

-   `summary_name = function(column)`: Name for the summary and how to calculate it
-   Common summary functions: `mean()`, `median()`, `sd()`, `var()`, `min()`, `max()`, `sum()`, `n()` (count)
-   Use `na.rm = TRUE` inside functions to handle missing values

```{r}
# Calculate summary statistics
summarize(car_data,
          mean_mpg = mean(mpg),
          median_mpg = median(mpg),
          sd_mpg = sd(mpg),
          n = n())
```

### `group_by()`: Group Operations

The `group_by()` function groups the data frame by one or more variables. Subsequent operations (especially `summarize()` and `mutate()`) are then performed separately for each group. This is similar to `GROUP BY` in SQL.

**Syntax:**

```{r, eval=FALSE}
group_by(data, grouping_column1, grouping_column2, ...)
```

**Key arguments:**

-   Column names to group by (unquoted)
-   Multiple columns create nested groups
-   Use `ungroup()` to remove grouping
-   Use `.groups = "drop"` in `summarize()` to automatically ungroup

The real power of `summarize()` comes when combined with `group_by()`:

```{r}
# Summary statistics by group
grouped_data = group_by(car_data, cyl)
summarize(grouped_data,
          mean_mpg = mean(mpg),
          median_mpg = median(mpg),
          sd_mpg = sd(mpg),
          count = n())
```

```{r}
# Group by multiple variables
grouped_data = group_by(car_data, cyl, am)
summarize(grouped_data,
          mean_mpg = mean(mpg),
          count = n(),
          .groups = "drop")
```

### `pivot_longer()`: Reshape from Wide to Long

The `pivot_longer()` function (from the `tidyr` package, also part of tidyverse) transforms data from wide format to long format. This means taking multiple columns and collapsing them into two columns: one for the variable names and one for the values.

**Syntax:**

```{r, eval=FALSE}
pivot_longer(data, cols, names_to = "name", values_to = "value")
```

**Key arguments:**

-   `cols`: Columns to pivot (can use helper functions like `starts_with()`, or ranges like `col1:col5`)
-   `names_to`: Name of the new column that will contain the original column names
-   `values_to`: Name of the new column that will contain the values

```{r}
# Create example wide data
wide_data = data.frame(
  car = c("Mazda RX4", "Datsun 710", "Hornet 4 Drive"),
  mpg = c(21.0, 22.8, 21.4),
  hp = c(110, 93, 110),
  wt = c(2.62, 2.32, 3.21)
)
wide_data
```

```{r}
# Pivot from wide to long format
long_data = pivot_longer(wide_data,
                          cols = c(mpg, hp, wt),
                          names_to = "measurement",
                          values_to = "value")
long_data
```

### `pivot_wider()`: Reshape from Long to Wide

The `pivot_wider()` function (from the `tidyr` package) transforms data from long format to wide format. This means spreading values from one column across multiple new columns.

**Syntax:**

```{r, eval=FALSE}
pivot_wider(data, names_from = "name_column", values_from = "value_column")
```

**Key arguments:**

-   `names_from`: Column whose values will become new column names
-   `values_from`: Column whose values will fill the new columns
-   `values_fill`: Value to use for missing combinations (optional)

```{r}
# Pivot from long back to wide format
pivot_wider(long_data,
            names_from = measurement,
            values_from = value)
```

## The Pipe Operator: `%>%`

The pipe operator `%>%` takes the output of one function and passes it as the first argument to the next function. This allows you to chain operations together in a readable way, reading left to right instead of inside out.

```{r}
# Without pipe: nested functions (read inside out)
head(arrange(car_data, mpg), 5)

# With pipe: read left to right
car_data %>% 
  arrange(mpg) %>% 
  head(5)
```

## Combining Multiple Operations

The pipe operator allows us to chain multiple operations together into a data manipulation pipeline:

```{r}
# Complex data manipulation pipeline
car_data %>% 
  filter(hp > 100) %>%
  select(X, mpg, hp, wt, cyl) %>%
  mutate(power_to_weight = hp / wt) %>%
  arrange(desc(power_to_weight)) %>%
  head(10)
```

```{r}
# Summary statistics by group using pipes
car_data %>% 
  group_by(cyl) %>% 
  summarize(
    mean_mpg = mean(mpg),
    median_mpg = median(mpg),
    sd_mpg = sd(mpg),
    count = n()
  )
```

## Useful Helper Functions

```{r}
# Count observations by group
count(car_data, cyl)
```

```{r}
# Rename columns
head(rename(car_data, miles_per_gallon = mpg, horsepower = hp), 3)
```

```{r}
# Get distinct values
distinct(car_data, cyl)
```

## Summary

In this lecture, we covered:

-   **dplyr and tidyr**: Using core verbs (`select()`, `filter()`, `arrange()`, `mutate()`, `summarize()`, `group_by()`, `pivot_longer()`, `pivot_wider()`) and the pipe operator for chaining operations

## References

-   Wickham, H., & Grolemund, G. (2017). *R for Data Science*. O'Reilly Media. Available free online at <https://r4ds.had.co.nz/>
-   dplyr documentation: <https://dplyr.tidyverse.org/>
-   tidyr documentation: <https://tidyr.tidyverse.org/>

```{r cleanup, include=FALSE}
if (file.exists("mtcars_example.csv")) {
  file.remove("mtcars_example.csv")
}
```
