# Data Structures in R Overview
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A *data structure* is a particular way of organizing data in a computer so that it can be used effectively. The idea is to reduce the space and time complexities of different tasks. Data structures in R programming are tools for holding multiple values.

R has **six** primary built-in data structures, which are broadly categorized by their dimensionality and whether they can store a single (homogeneous) or multiple (heterogeneous) data types.

## Vector

Vectors are the simplest form of data structure in R. They are a collection of elements of the same type, such as numeric, character, or logical.

A vector is created by `c()` command, short for *combine*:

```{r}
x = c(1,2,3)
print(x)
y = c("apple","orange","banana")
print(y)
```

### Vectorized Operations

You can apply a function to a vector without needing to loop through its elements explicitly. For example, adding two vectors together adds corresponding elements in a faster and more concise way than doing so element-wise through loops.

```{r}
x = c(2,4,6)
y = c(7,5,3)
x+y # add vector x to vector y by elements
x-y # subtract elements in vector y from corresponding elements in vector x
x*y # multiply vector x by vector y by elements
```

[***Question 1:***]{.underline} What will the following codes return?

```{r eval=FALSE, include=TRUE}
x = c(2,4,6)
y = c(7,5,3)
x < y
x == y
x > y
x == 3
```

### Accessing Vector Elements

Elements in a vector are accessed using square brackets `[]`. R uses 1-based indexing, meaning the first element is at index 1, not 0.

```{r}
example = c(10, 20, 30, 40, 50)
# Extract the third element
example[3] 
# Extract elements from the second to fourth position
example[2:4] 
# Extract non-sequential elements, e.g. the first element and the fourth element
example[c(1, 4)] 
```

### Remove Vector Elements

Specific elements in a vector can be excluded by using negative indices.

```{r}
example = c(10, 20, 30, 40, 50)
# Remove the third element
example[-3] 
# Remove elements from the second to fourth position
example[-(2:4)] 
# Remove non-sequential elements, e.g. the first element and the fourth element
example[-c(1, 4)] 
```

### Coercion in a Vector

All elements of a vector must be the same type, so when you attempt to combine different types they will be coerced to the most flexible type. Types from least to most flexible are: logical, integer, double, and character. For instance, combining a character and an integer yields a character.

## Lists

Lists are versatile data structures in R that can hold a mix of objects of different types and sizes, even other lists. You can create a list using the `list()` function.

```{r}
x = list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
print(x)
```

You can also name objects included in a list:

```{r}
anotherlist = list(name="Tina Belcher", year=2025, scores=c(80, 90, 85), female =TRUE)
print(anotherlist)
```

Once you have created the list, you can access its elements using either indices or the name.

```{r}
print(anotherlist[[3]])
print(anotherlist$name)
print(anotherlist$scores)
```

[***Question 2:***]{.underline} Extract the second score (i.e. 90) from `anotherlist`.

## Matrices

A matrix is a two-dimensional data structure where all elements must be of the same type (homogeneous). You can create a matrix using the `matrix()` function. If no specifications are explicitly stated, by default, matrices are filled column-wise.

```{r}
# Create a 3x3 matrix filled by column (default)
m = matrix(1:9, nrow = 3, ncol = 3)
print(m)
```

To fill row-wise, use the `byrow` argument:

```{r}
# Create a 3x3 matrix filled by row
m_byrow = matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
print(m_byrow)
```

### Naming Rows and Columns

You can assign names to rows and columns using `rownames()` and `colnames()`:

```{r}
grades = matrix(c(85, 90, 78, 92, 88, 95), nrow = 2, ncol = 3, byrow = TRUE)
rownames(grades) = c("Alice", "Bob")
colnames(grades) = c("Exam1", "Exam2", "Exam3")
print(grades)
```

### Accessing Matrix Elements

Matrix elements are accessed using `[row, column]` notation. Note that, by default, subsetting a single row or column from a matrix returns a vector, not a column- or row-matrix as might be expected.

```{r}
# Access element in row 1, column 2
grades[1, 2]
# Access entire first row
grades[1, ]
# Access entire second column
grades[, 2]
# Access a submatrix
grades[1:2, 2:3]
```

### Matrix Operations

R supports element-wise and matrix algebra operations:

```{r}
A = matrix(1:4, nrow = 2)
B = matrix(5:8, nrow = 2)

# Element-wise multiplication
A * B

# Matrix multiplication
A %*% B

# Transpose
t(A)
```

[***Question 3:***]{.underline} Create a 2x4 matrix containing the numbers 10 through 17, filled by row. Then extract the element in the second row, third column.

## Arrays

Arrays extend matrices to more than two dimensions. All elements must still be of the same type. You create an array using the `array()` function. R populates arrays layer by layer, filling each column from top to bottom before moving to the next column, and completing each layer before advancing to the next. In the example below, the first nine values fill the first layer column-wise, and the remaining nine fill the second.

```{r}
# Create a 3x3x2 array (3 rows, 3 columns, 2 "layers")
arr = array(1:18, dim = c(3, 3, 2))
print(arr)
```

### Accessing Array Elements

Array elements are accessed by specifying indices for each dimension:

```{r}
# Access element at row 2, column 3, layer 1
arr[2, 3, 1]
# Access the entire first layer
arr[, , 1]
# Access the second row across all layers
arr[2, , ]
```

### Naming Array Dimensions

You can name each dimension using the `dimnames` argument:

```{r}
arr_named = array(
  1:12, 
  dim = c(2, 3, 2),
  dimnames = list(
    c("Row1", "Row2"),
    c("Col1", "Col2", "Col3"),
    c("Layer1", "Layer2")
  )
)
print(arr_named)
```

[***Question 4:***]{.underline} Create a 2x2x3 array. How many total elements does it contain?

## Data Frames

Data from a study or an experiment often consist of different types of measurements on a set of subjects or experimental units. For example, the World Bank provides summary statistics on countries around the world. In this case, the unit is the country and the measurements include gross domestic product, life expectancy, and literacy rate. These data are naturally organized into a table format with one row for each country and one column for each summary statistic.

The data frame is a data structure designed for the purpose of working with data that naturally organize into rows that correspond to observations/units/individuals and columns that correspond to measurements take on the observations.

Data frames are the most commonly used data structure for data analysis in R. They are two-dimensional, like matrices, but can hold columns of different types (heterogeneous).

To create an R data frame use `data.frame()` function and then pass each of the vectors we have created as arguments to the function:

```{r}
# Create a data frame
students = data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 22, 21),
  major = c("Math", "Statistics", "Physics"),
  gpa = c(3.8, 3.5, 3.9)
)
print(students)
```

### Examining Data Frame Structure

Several functions help you understand your data frame:

```{r}
# View structure
str(students)
# View first few rows
head(students)
# Get dimensions
dim(students)
# Get number of rows and columns
nrow(students)
ncol(students)
# Get column names
names(students)
```

### Accessing Data Frame Elements

You can access data frame elements using indices, column names, or the `$` operator:

```{r}
# Access by column name
students$name
# Access by column index
students[, 2]
# Access by row and column
students[1, 3]
# Access multiple columns
students[, c("name", "gpa")]
# Access rows based on condition
students[students$gpa > 3.6, ]
```

### Adding and Modifying Columns

```{r}
# Add a new column
students$enrolled = c(TRUE, TRUE, FALSE)
print(students)
# Modify an existing column
students$age = students$age + 1
print(students)
```

### Useful Data Frame Functions

```{r}
# Summary statistics for each column
summary(students)
```

[***Question 5:***]{.underline} Using the `students` data frame, extract all rows where the student's major is "Math" or "Statistics".

## Factors

Factors are used to represent categorical data in R. They store both the values and the predefined levels (categories) the data can take. Factors are essential for statistical modeling, as they tell R to treat the variable as categorical rather than continuous.

```{r}
# Create a factor
sizes = factor(c("Small", "Medium", "Large", "Medium", "Small"))
print(sizes)
# View the levels
levels(sizes)
```

### Ordered Factors

For ordinal data (categories with a meaningful order), use ordered factors:

```{r}
# Create an ordered factor
sizes_ordered = factor(
  c("Small", "Medium", "Large", "Medium", "Small"),
  levels = c("Small", "Medium", "Large"),
  ordered = TRUE
)
print(sizes_ordered)
# Comparisons work with ordered factors
sizes_ordered[1] < sizes_ordered[3]
```

### Changing Factor Levels

You can rename or reorder levels:

```{r}
# Reorder levels
education = factor(c("High School", "Bachelor", "Master", "Bachelor"))
education = factor(education, levels = c("High School", "Bachelor", "Master", "PhD"))
print(education)
levels(education)
```

### Factors in Data Frames

When creating data frames, character columns are NOT automatically converted to factors in modern R (version 4.0+). You can explicitly convert them:

```{r}
students$major = factor(students$major)
str(students)
```

[***Question 6:***]{.underline} Create a factor variable representing letter grades (A, B, C, D, F) for five students. Make it an ordered factor with F as the lowest and A as the highest.

# Summary Table

| Data Structure | Dimensions | Homogeneous? | Creation Function |
|----------------|------------|--------------|-------------------|
| Vector         | 1D         | Yes          | `c()`             |
| List           | 1D         | No           | `list()`          |
| Matrix         | 2D         | Yes          | `matrix()`        |
| Array          | nD         | Yes          | `array()`         |
| Data Frame     | 2D         | No           | `data.frame()`    |
| Factor         | 1D         | Yes          | `factor()`        |
